### 2025-07-03 ms/ms.tex
updated ms

\textsuperscript{1} CAS Key Laboratory of Tropical Forest Ecology,
Xishuangbanna Tropical Botanical Garden, Chinese Academy of Sciences,
Menglun, Yunnan 666303, China

@@

where \(f(\cdot)\) predicts the expected natural log tree height or
crown radius from observed DBH (and \emph{vice versa}),
\(\pmb{\beta}_{j}=\{a_j, b_j, k_j\}\) denotes the vector of
species-specific parameters, and \(\sigma_y\) represents the standard
deviation of the observation error. We considered three functional forms
for the relationship between tree allometric variables
(\citeproc{ref-Sullivan2018}{Sullivan et al. 2018};
\citeproc{ref-MartinezCano2019}{Martínez Cano et al. 2019}):

@@

We applied the power-law model to estimate tree height and crown radius
from DBH, as well as DBH from tree height and/or crown radius.
Specifically, when predicting DBH from height and/or crown radius, we
tested four formulations: (1) using CR and H as separate factors,
\(\mathrm{DBH} = a_1 \mathrm{CR}^{b_1} \mathrm{H}^{c_1}\), denoted as
\(CR,H\); (2) using CR and H as a compound term,
\(\mathrm{DBH} = a_2 (\mathrm{CR} \times \mathrm{H})^{b_2}\), denoted as
\(CR \times H\); (3) using only height,
\(\mathrm{DBH} = a_3 \mathrm{H}^{b_3}\); and (4) using only crown
radius, \(\mathrm{DBH} = a_4 \mathrm{CR}^{b_4}\). In contrast, the gMM
and Weibull models were only used to estimate tree height and crown
radius from DBH. Because these two models specifically describe
saturating relationships (Eq.~\ref{eq-gMM}; Eq.~\ref{eq-wb}), their
formulation does not allow simply switching the roles of explanatory and
response variables.

To compare the performance of different models for the same response
variable, we used Pareto-smoothed importance sampling leave-one-out
cross-validation (PSIS-LOO; Vehtari et al.
(\citeproc{ref-Vehtari2017}{2017})). PSIS-LOO is an accurate and
reliable approximation to standard leave-one-out cross-validation (LOO),
which is a robust method for comparing models with different numbers of
parameters (\citeproc{ref-Vehtari2017}{Vehtari et al. 2017}). We
utilized PSIS-LOO to calculate the LOO Information Criterion (LOOIC) for
each model. A model with lower LOOIC is considered to have better
predictive accuracy (\citeproc{ref-Vehtari2017}{Vehtari et al. 2017}).

@@

where \(\gamma_{0}\) and \(\gamma_{1}\) are the species-level intercept
and slope, respectively, and \(\sigma_{\beta}\) is the standard
deviation of the species-specific parameters. To facilitate
interpretation and visualization, species were then classified into
``light'' and ``dense'' wood groups based on the 10th and 90th
percentiles of wood density distribution in the observed dataset,
respectively. This classification allowed us to illustrate the
relationships between wood density and the parameters of the
best-performing models, enabling an evaluation of how wood density
contributes to variation in tree architecture. Additionally, we explored
a model that accounts for covariance between species-specific parameters
using a multivariate normal distribution. However, due to the complexity
of this approach, the model did not converge for the nonlinear models,
and thus, we did not include it in the final analysis.

Posterior distributions of all parameters were estimated using the
No-U-Turn Sampler (NUTS), an adaptive variant of the Hamiltonian Monte
Carlo (HMC), algorithm implemented in Stan
(\citeproc{ref-Carpenter2017}{Carpenter et al. 2017}) using
weakly-informative priors (\citeproc{ref-Gelman2008}{Gelman et al.
2008}). The HMC algorithm uses gradient information to propose new
states in the Markov chain, leading to a more efficient exploration of
the target distribution than traditional Markov chain Monte Carlo (MCMC)
methods that rely on random proposals
(\citeproc{ref-Carpenter2017}{Carpenter et al. 2017}). This efficiency
allows us to achieve convergence with fewer iterations than traditional
MCMC methods. Inferences were based on 1,000 posterior samples following
1,000 warmup iterations for three parallel chains. Convergence was
assessed using the Gelman--Rubin statistic with a convergence threshold
of 1.1 (\citeproc{ref-Gelman2013}{Gelman et al. 2015}) and ensuring
effective sample sizes greater than 300
(\citeproc{ref-Vehtari2021}{Vehtari et al. 2021}). We also reported the
median and 95\% Bayesian credible intervals (CIs) for each parameter
(Supplementary Tables S3-S8). Estimates were considered significant when
95\% CIs did not overlap with zero.

@@

In CR-DBH allometry, angiosperms exhibited a positive relationship
between the scaling factor \(a\) and wood density and a negative
association between the exponent parameter \(b\) and wood density
(Fig.~\ref{fig-wood1} g, h; Fig.~\ref{fig-wood2} b). In gymnosperms, we
found a positive association between wood density and the exponent
parameter \(b\), whereas neither the asymptote parameter \(a\) nor the
scaling constant \(k\) showed significant relationship with this
functional trait (Fig.~\ref{fig-wood1} i, j, k; Fig.~\ref{fig-wood2} c).

@@

AGB estimates based on the Weibull model (\(\text{AGB}_\text{H-WB}\))
showed the highest predictive accuracy based both on the lowest root
mean square error (RMSE) of 0.279 Mg and the smallest bias of 2.04\%
(Fig.~\ref{fig-agb} a). Estimates based on the power-law model
(\(\text{AGB}_\text{H-PL}\)) also showed competitive performance but had
higher RMSE and bias compared to the Weibull-based approach, with RMSE
of 0.28 Mg and bias of 2.29\% (Fig.~\ref{fig-agb} b). In contrast,
models that used estimated DBH performed substantially worse
(Fig.~\ref{fig-agb} c-f).

@@

Our result supported a saturating H-DBH relationship in both angiosperm
and gymnosperm species, consistent with theoretical expectations
(\citeproc{ref-Falster2003}{Falster and Westoby 2003};
\citeproc{ref-Niklas2007}{Niklas 2007}) and empirical observations
(e.g., \citeproc{ref-Feldpausch2011}{Feldpausch et al. 2011};
\citeproc{ref-MartinezCano2019}{Martínez Cano et al. 2019};
\citeproc{ref-Song2024}{Song et al. 2024}). A recent study using Tallo
suggested the Weibull function as the best model for six plant
functional types (PFTs) (\citeproc{ref-Song2024}{Song et al. 2024}).
While PFTs-based approaches can illuminate large-scale patterns, they
may obscure species-specific variation in allometric scaling, and
locally abundant species may bias the overall estimates. Our best
predictive model effectively captures species-specific patterns
(Supplementary Table S6) despite underestimating the height of
individuals with DBH \textgreater{} 30 cm due to substantial
interspecific variation in H--DBH allometric scaling
(Fig.~\ref{fig-compare}). This highlights the advantages of a multilevel
modeling approach in addressing systematic differences across species.
Additionally, the metabolic scaling theory (MST) posits logarithmic
scaling relationships among different axes of tree size, with an
exponent close to 2/3 (\(H \propto DBH^{2/3}\),
\(CR \propto DBH^{2/3}\)), based on assumptions regarding mechanical and
hydraulic constraints to plant growth (\citeproc{ref-West1999}{West et
al. 1999}). However, our findings revealed that the power-law model
exponent of angiosperms (\emph{b} = 0.538) was significantly smaller,
while the actual exponent in gymnosperms (\emph{b} = 0.704) was larger
than the theoretical value of 2/3 (Supplementary Table S3). This
divergence highlights the importance of separately modeling the
allometric relationships of each functional group to accurately capture
their distinct scaling patterns. Moreover, the power-law function
performed worse than the saturating functions (Table 1), indicating that
the actual H-DBH relationship does not follow the simple scaling laws
proposed by MST.

Our analysis also revealed divergent patterns in CR-DBH allometry,
highlighting fundamental differences in growth strategies and ecological
adaptations between angiosperms and gymnosperms. In angiosperms, the
CR--DBH relationship follows a power-law form, suggesting a capacity for
plastic crown expansion in competitive environments
(\citeproc{ref-Lines2012}{Lines et al. 2012}; e.g.,
\citeproc{ref-Poorter2012a}{Poorter et al. 2012a};
\citeproc{ref-Hulshof2015}{Hulshof et al. 2015}), a conclusion
consistent with a study in tropical moist forests by Martínez Cano et
al. (\citeproc{ref-MartinezCano2019}{2019}). In addition, the estimated
exponent aligns closely with MST predictions for crown radius,
supporting the selection of the power-law model for angiosperms, where
the community-level exponent \emph{b} is near the expected 2/3
(\(b_{\text{ang}}\) = 0.626 {[}0.617, 0.635{]}, Table 2). In contrast,
the CR-DBH allometry in gymnosperms showed an asymptotic relationship
best described by the gMM function, with the Weibull function also
showing competitive predictive performance (Table 1). This asymptotic
pattern likely arises from structural constraints in gymnosperms growth
forms that favour vertical over lateral expansion (e.g.,
\citeproc{ref-Lines2012}{Lines et al. 2012};
\citeproc{ref-Hulshof2015}{Hulshof et al. 2015};
\citeproc{ref-Jucker2024}{Jucker et al. 2024}), an adaptation that can
mitigate risks such as branch breakage under heavy snow loads (e.g.,
\citeproc{ref-Mikola1938}{Mikola 1938}; \citeproc{ref-Loehle2016}{Loehle
2016}; \citeproc{ref-Tavankar2019}{Tavankar et al. 2019}).

@@

Our analysis indicated distinct patterns in which wood density has a
greater influence on the interspecific variation in angiosperms than
gymnosperms. Variation in wood density has consequential implications
for the mechanical properties of wood and hence influences the
mechanical stability of stems and branches
(\citeproc{ref-Sterck1998}{Sterck and Bongers 1998}), ultimately
determining the architecture of trees (e.g.,
\citeproc{ref-Poorter2006}{Poorter et al. 2006}). Angiosperms have
denser wood and a wider wood density range than gymnosperms (see
Supplementary Table S2), resulting in greater and more flexible
mechanical support (e.g., \citeproc{ref-Gelder2006}{Gelder et al. 2006};
\citeproc{ref-Swenson2007}{Swenson and Enquist 2007};
\citeproc{ref-Mo2024}{Mo et al. 2024}). This allows them to develop a
more diverse height and crown to optimize light capture, particularly in
environments where light competition is intense, such as tropical
regions. In contrast, gymnosperms predominantly live and often dominate
in extreme environments such as boreal and high-elevation regions
(\citeproc{ref-Swenson2007}{Swenson and Enquist 2007};
\citeproc{ref-Mo2024}{Mo et al. 2024}), prioritising survival and
longevity (\citeproc{ref-Fossdal2024}{Fossdal et al. 2024}). Due to
their simpler and more homogeneous wood anatomy, these species display
lighter wood and a narrower range of wood density (e.g.,
\citeproc{ref-Mo2024}{Mo et al. 2024}), which may result in less
flexibility in the crown architecture. These differences underscore how
the contrasting evolutionary strategies of angiosperms and gymnosperms
shape their architectural diversity, reflecting adaptations to distinct
ecological niches.

>>>>

\textsuperscript{1} Yunnan Key Laboratory of Forest Ecosystem Stability
and Global Change Response, Xishuangbanna Tropical Botanical Garden,
Chinese Academy of Sciences, Menglun, Yunnan 666303, China

@@

where \(f(\cdot)\) predicts the expected natural log tree height or
crown radius from observed DBH (and \emph{vice versa}),
\(\pmb{\beta}_{j}=\{a_j, b_j, c_j, k_j\}\) denotes the vector of
species-specific parameters, and \(\sigma_y\) represents the standard
deviation of the observation error. We considered three functional forms
for the relationship between tree allometric variables
(\citeproc{ref-Sullivan2018}{Sullivan et al. 2018};
\citeproc{ref-MartinezCano2019}{Martínez Cano et al. 2019}):

@@

We applied the power-law model to estimate tree height and crown radius
from DBH, as well as DBH from tree height and/or crown radius.
Specifically, when predicting DBH from height and/or crown radius, we
tested four formulations: (1) using CR and H as separate factors,
\(\mathrm{DBH} = a_1 \mathrm{CR}^{b_1} \mathrm{H}^{c_1}\), denoted as
\(CR,H\); (2) using CR and H as a compound term,
\(\mathrm{DBH} = a_2 (\mathrm{CR} \times \mathrm{H})^{b_2}\), denoted as
\(CR \times H\); (3) using only height,
\(\mathrm{DBH} = a_3 \mathrm{H}^{b_3}\); and (4) using only crown
radius, \(\mathrm{DBH} = a_4 \mathrm{CR}^{b_4}\). In contrast, the gMM
and Weibull models were only used to estimate tree height and crown
radius from DBH. Because these two models specifically describe
saturating relationships (Eq.~\ref{eq-gMM}; Eq.~\ref{eq-wb}), their
formulation does not allow for simply switching the roles of explanatory
and response variables.

To compare the performance of different models for the same response
variable, we used Pareto-smoothed importance sampling leave-one-out
cross-validation (PSIS-LOO; Vehtari et al.
(\citeproc{ref-Vehtari2017}{2017})). PSIS-LOO is an accurate and
reliable approximation to standard leave-one-out cross-validation (LOO),
a robust method for comparing models with different numbers of
parameters (\citeproc{ref-Vehtari2017}{Vehtari et al. 2017}). We
utilized PSIS-LOO to calculate the LOO Information Criterion (LOOIC) for
each model. A model with a lower LOOIC is considered to have better
predictive accuracy (\citeproc{ref-Vehtari2017}{Vehtari et al. 2017}).

@@

where \(\gamma_{0}\) and \(\gamma_{1}\) are the global-level
hyperparameters for intercept and slope, respectively, and
\(\sigma_{\beta}\) is the standard deviation of the species-specific
parameters. To facilitate interpretation and visualization, species were
then classified into ``light'' and ``dense'' wood groups based on the
10th and 90th percentiles of wood density distribution in the observed
dataset, respectively. This classification allowed us to illustrate the
relationships between wood density and the parameters of the
best-performing models, enabling an evaluation of how wood density
contributes to variation in tree architecture. Additionally, we explored
a model that accounts for covariance between species-specific parameters
using a multivariate normal distribution. However, due to the complexity
of this approach, the model did not converge for the nonlinear models,
and thus, we did not include it in the final analysis.

Posterior distributions of all parameters were estimated using the
No-U-Turn Sampler (NUTS), an adaptive variant of the Hamiltonian Monte
Carlo (HMC), algorithm implemented in Stan
(\citeproc{ref-Carpenter2017}{Carpenter et al. 2017}) using
weakly-informative priors (\citeproc{ref-Gelman2008}{Gelman et al.
2008}). The HMC algorithm uses gradient information to propose new
states in the Markov chain, leading to a more efficient exploration of
the target distribution than traditional Markov chain Monte Carlo (MCMC)
methods that rely on random proposals
(\citeproc{ref-Carpenter2017}{Carpenter et al. 2017}). This efficiency
allows us to achieve convergence with fewer iterations than traditional
MCMC methods. Inferences were based on 1,000 posterior samples following
1,000 warmup iterations for three parallel chains. Convergence was
assessed using the Gelman--Rubin statistic with a convergence threshold
of 1.1 (\citeproc{ref-Gelman2013}{Gelman et al. 2015}) and ensuring
effective sample sizes greater than 300
(\citeproc{ref-Vehtari2021}{Vehtari et al. 2021}). We also reported the
median and 95\% Bayesian credible intervals (CIs) for each parameter
(Supplementary Tables S3-S8). Estimates were considered significant when
their 95\% CIs did not overlap with zero.

@@

In CR-DBH allometry, angiosperms exhibited a positive relationship
between the scaling factor \(a\) and wood density and a negative
association between the exponent parameter \(b\) and wood density
(Fig.~\ref{fig-wood1} g, h; Fig.~\ref{fig-wood2} b). In gymnosperms, we
found a positive association between wood density and the exponent
parameter \(b\), whereas neither the asymptote parameter \(a\) nor the
scaling constant \(k\) showed a significant relationship with this
functional trait (Fig.~\ref{fig-wood1} i, j, k; Fig.~\ref{fig-wood2} c).

@@

AGB estimates based on the Weibull model (\(\text{AGB}_\text{H-WB}\))
showed the highest predictive accuracy based on both the lowest root
mean square error (RMSE) of 0.279 Mg and the smallest bias of 2.04\%
(Fig.~\ref{fig-agb} a). Estimates based on the power-law model
(\(\text{AGB}_\text{H-PL}\)) also showed competitive performance but had
higher RMSE and bias compared to the Weibull-based approach, with RMSE
of 0.28 Mg and bias of 2.29\% (Fig.~\ref{fig-agb} b). In contrast,
models that used estimated DBH performed substantially worse
(Fig.~\ref{fig-agb} c-f).

@@

Our results supported a saturating H-DBH relationship in both angiosperm
and gymnosperm species, consistent with theoretical expectations
(\citeproc{ref-Falster2003}{Falster and Westoby 2003};
\citeproc{ref-Niklas2007}{Niklas 2007}) and empirical observations
(e.g., \citeproc{ref-Feldpausch2011}{Feldpausch et al. 2011};
\citeproc{ref-MartinezCano2019}{Martínez Cano et al. 2019};
\citeproc{ref-Song2024}{Song et al. 2024}). A recent study using Tallo
suggested the Weibull function as the best model for six plant
functional types (PFTs) (\citeproc{ref-Song2024}{Song et al. 2024}).
While PFTs-based approaches can illuminate large-scale patterns, they
may obscure species-specific variation in allometric scaling, and
locally abundant species may bias the overall estimates. Our best
predictive model effectively captures species-specific patterns
(Supplementary Table S6) despite underestimating the height of
individuals with DBH \textgreater{} 30 cm due to substantial
interspecific variation in H--DBH allometric scaling
(Fig.~\ref{fig-compare}). This highlights the advantages of a multilevel
modeling approach in addressing systematic differences across species.
Additionally, the metabolic scaling theory (MST) posits logarithmic
scaling relationships among different axes of tree size, with an
exponent close to 2/3 (\(H \propto DBH^{2/3}\),
\(CR \propto DBH^{2/3}\)), based on assumptions regarding mechanical and
hydraulic constraints to plant growth (\citeproc{ref-West1999}{West et
al. 1999}). However, our findings revealed that the power-law model
exponent of angiosperms (\emph{b} = 0.538) was significantly smaller,
while the actual exponent in gymnosperms (\emph{b} = 0.704) was larger
than the theoretical value of 2/3 (Supplementary Table S3). This
divergence highlights the importance of separately modeling the
allometric relationships of each functional group to accurately capture
their distinct scaling patterns. Moreover, the power-law function
performed worse than the saturating functions (Table 1), indicating that
the actual H-DBH relationship does not follow the simple scaling laws
proposed by MST.

Our analysis also revealed divergent patterns in CR-DBH allometry,
highlighting fundamental differences in growth strategies and ecological
adaptations between angiosperms and gymnosperms. In angiosperms, the
CR--DBH relationship follows a power-law form, suggesting a capacity for
plastic crown expansion in competitive environments
(\citeproc{ref-Lines2012}{Lines et al. 2012}; e.g.,
\citeproc{ref-Poorter2012a}{Poorter et al. 2012a};
\citeproc{ref-Hulshof2015}{Hulshof et al. 2015}), a conclusion
consistent with a study in tropical moist forests by Martínez Cano et
al. (\citeproc{ref-MartinezCano2019}{2019}). In addition, the estimated
exponent aligns closely with MST predictions for crown radius,
supporting the selection of the power-law model for angiosperms, where
the community-level exponent \emph{b} is near the expected 2/3
(\(b_{\text{ang}}\) = 0.626 {[}0.617, 0.635{]}, Table 2). In contrast,
the CR-DBH allometry in gymnosperms showed an asymptotic relationship
best described by the gMM function, with the Weibull function also
showing competitive predictive performance (Table 1). This asymptotic
pattern likely arises from structural constraints in gymnosperms' growth
forms that favor vertical over lateral expansion (e.g.,
\citeproc{ref-Lines2012}{Lines et al. 2012};
\citeproc{ref-Hulshof2015}{Hulshof et al. 2015};
\citeproc{ref-Jucker2024}{Jucker et al. 2024}), an adaptation that can
mitigate risks such as branch breakage under heavy snow loads (e.g.,
\citeproc{ref-Mikola1938}{Mikola 1938}; \citeproc{ref-Loehle2016}{Loehle
2016}; \citeproc{ref-Tavankar2019}{Tavankar et al. 2019}).

@@

Our analysis indicated distinct patterns in which wood density has a
greater influence on interspecific variation in angiosperms than in
gymnosperms. Variation in wood density has consequential implications
for the mechanical properties of wood and hence influences the
mechanical stability of stems and branches
(\citeproc{ref-Sterck1998}{Sterck and Bongers 1998}), ultimately
determining the architecture of trees (e.g.,
\citeproc{ref-Poorter2006}{Poorter et al. 2006}). Angiosperms have
denser wood and a wider wood density range than gymnosperms (see
Supplementary Table S2), resulting in greater and more flexible
mechanical support (e.g., \citeproc{ref-Gelder2006}{Gelder et al. 2006};
\citeproc{ref-Swenson2007}{Swenson and Enquist 2007};
\citeproc{ref-Mo2024}{Mo et al. 2024}). This allows them to develop a
more diverse height and crown to optimize light capture, particularly in
environments where light competition is intense, such as tropical
regions. In contrast, gymnosperms predominantly live and often dominate
in extreme environments such as boreal and high-elevation regions
(\citeproc{ref-Swenson2007}{Swenson and Enquist 2007};
\citeproc{ref-Mo2024}{Mo et al. 2024}), prioritising survival and
longevity (\citeproc{ref-Fossdal2024}{Fossdal et al. 2024}). Due to
their simpler and more homogeneous wood anatomy, these species display
lighter wood and a narrower range of wood density (e.g.,
\citeproc{ref-Mo2024}{Mo et al. 2024}), which may result in less
flexibility in the crown architecture. These differences underscore how
the contrasting evolutionary strategies of angiosperms and gymnosperms
shape their architectural diversity, reflecting adaptations to distinct
ecological niches.
